%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                                                      %
%     File: Thesis_Abstract.tex                                        %
%     Tex Master: Thesis.tex                                           %
%                                                                      %
%     Author: Andre C. Marta                                           %
%     Last modified :  2 Jul 2015                                      %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{Abstract}

% Add entry in the table of contents as section
\addcontentsline{toc}{section}{Abstract}

This thesis presents a new simulation environment for the RV32-Versat system
using the Verilator simulation framework. The RV32-Versat system consists of the
PicoRV32 RISC-V processor connected to the Versat \ac{CGRA} architecture. This new 
simulation environment presents significant advantages over
the typical simulation environments using commercial event-driven
simulators. The first advantage is speed: the new environment is considerably
faster, therefore saving valuable time when developing applications for the
RV32-Versat architecture. The second advantage is cost: Verilator is available
under a free open-source licence, while the typical event-driven simulators
require very expensive licences, hard to justify for small companies and
projects. The third and last advantage is the easy support for hardware and
software co-simulation. The event-driven simulators are lacking in this field,
but the new environment using Verilator solves this problem, allowing a seamless
integration through C++ or SystemC. This new simulation environment is the
result of a detailed study of the state of the art of \ac{CPU} and \ac{CGRA}
simulation and the RV32-Versat architecture, also presented in this thesis.

\vfill

\textbf{\Large Keywords:}  Coarse-Grain Reconfigurable Arrays, Simulation Environment, 
Verilator, CGRA Simulation, High-Level Simulation, 
Co-Simulation


